Sparse spectral and $p$-finite element methods for partial differential equations on disk-slices and trapeziums

### 2019-04-25 papers/halfdisk/halfdisk.tex
Minor edit
Readthrough

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials. 
\end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align}
where we require that either of the following scenarios hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\).
\end{enumerate}

For simplicity of discussion we focus in this paper on the half-disk, where $\rho(x) = \sqrt{1-x^2}$,  $(\alpha,\beta) = (0, 1)$, and  $(\gamma, \delta)  = (-1,1)$, with discussion of extension to other geometries in the appendix.

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b
$$
with corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs on a half-sphere can be represented using two families of OPs on the half-disk \cite[Theorem 3.1]{olver2018orthogonal}, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

@@

We can generalise this for our purposes in the following proposition.
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
Then the four-parameter 2D polynomials defined on $\Omega$ given by
\begin{align}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
are orthogonal polynomials with respect to the weight 
\begin{align}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align}
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align}
Note here the we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align}

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$.

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align}
for \((x,y) \in \Omega\), where
\begin{align}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)} \\
\alphaab_{n,k,2} &:= \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align}

@@

\begin{proof}
First, note that, using a change of variable,
\begin{align}
	\norm{\hdopnkab}_\Wab^2 &= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}(\frac{y}{\rho(x)}}^2 \: \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(y)^2 \: w_P^{(b)}(y) \: \D y \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.

>>>>

\begin{abstract}
In recent years, sparse spectral methods for solving partial differential equations have been derived using hierarchies of classical orthogonal polynomials on intervals, disks, and triangles. In this work we extend this methodology to a hierarchy of non-classical orthogonal polynomials on disk slices (e.g. a half disk) and trapeziums. This builds on the observation that sparsity is guaranteed due to the boundary being defined by an algebraic curve, and that the entries of partial differential operators can be determined using formulae in terms of (non-classical) univariate orthogonal polynomials.  We apply the framework to solving the Poisson, variable coefficient Helmholtz, and Biharmonic equations.
 \end{abstract}

@@

This paper develops sparse spectral methods for solving linear partial differential equations on a special class of geometries that includes disk slices and trapeziums.  
More precisely, we consider the solution of partial differential equations on the domain
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
where  either of the following scenarios hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\).
\end{enumerate}

For simplicity of presentation we focus on the half-disk, where $\rho(x) = \sqrt{1-x^2}$,  $(\alpha,\beta) = (0, 1)$, and  $(\gamma, \delta)  = (-1,1)$, and discuss extension to other geometries in the appendix.

We show that partial differential equations become sparse linear systems when viewed as acting on expansions involving a family of orthogonal polynomials (OPs) that  generalise Jacobi polynomials, mirroring the ultraspherical spectral method for ordinary differential equations \cite{olver2013fast} and its analogue on the triangle \cite{olver2018recurrence,olver2019triangle}.  On the half-disk the family of weights we consider are of the form
$$
W^{(a,b)}(x,y) = x^a (1-x-y)^b, \qfor 0 \leq x \leq 1,\quad -\sqrt{1-x^2} \leq y \leq \sqrt{1-x^2},
$$
with corresponding OPs denoted $\hdopnkab(x,y)$, where $n$ denotes the polynomial degree, and $0 \le k \le n$. We define these to be orthogonalised lexicographically, that is,
$$
\hdopnkab(x,y) = C_{n,k} x^{n-k} y^k + (\hbox{lower order terms})
$$
where $C_{n,k} \neq 0$ and ``lower order terms'' can  include degree $n$ polynomials with fewer $y$ terms.

@@

The motivation for this work is solving partial differential equations on sub-domains of the sphere. In particular, OPs in cartesian coordinates ($x$, $y$, and $z$) on a half-sphere  can be represented using two families of OPs on the half-disk, see \cite[Theorem 3.1]{olver2018orthogonal} for a similar construction of OPs on an arc in 2D.
%, where we would require $2n+1$ 3D OPs for each degree $n$, of which $n+1$ would be the $n+1$ degree $n$ OPs on the half-disk of the first family (polynomials in $x,y$) while the remaining $n$ OPs would be given by $z$ multiplied by the $n$ degree $n-1$ OPs on the half-disk of the second family. 
Constructing sparse spectral methods for surface PDEs on half-spheres, spherical caps, and spherical triangles is future work, and has applications in weather prediction. Other extensions include a full $hp$-finite element method on sections of a disk, which has applications in fluids.

@@

\noindent\appref{PFEM}: We use the procedure to construct sparse $p$-finite element methods.

\noindent\appref{DiskSlices}: We discuss extension to disk slices.

\noindent\appref{Trapezium}: We discuss extension to trapezia.

@@

We can generalise\sotodo{specialise?} this for our purposes in the following proposition.
\sotodoinline{Can this be simplified to just the definition?}
\begin{proposition}
Let
\begin{align*}
	\Omega := \{(x,y) \in \R^2 \quad | \quad \alpha < x < \beta, \: \gamma \rho(x) < y < \delta \rho(x)\}
\end{align*}
with $\rho : (\alpha, \beta) \to \R$. Further, let $\genjacw^{(a,b,c)}(x)$ and $w^{(a,b)}_P(x)$ be two weight functions on the intervals $(\alpha, \beta)$ and $(\gamma, \delta)$ respectively, given by:
\begin{align*}
\begin{cases}
\genjacw^{(a,b,c)}(x) &:= (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{c} \\
w^{(a,b)}_P(x) &:= (\delta-x)^a \: (x - \gamma)^b
\end{cases}
\end{align*}
where the three-parameter family of orthogonal polynomials on $[\alpha,\beta]$ given by $\{\genjac_n^{(a,b,c)}\}$ are orthogonal with respect to $\genjacw^{(a,b,c)}(x)$, and the two-parameter family of orthogonal polynomials on $[\gamma,\delta]$ given by $\{P_n^{(a,b)}\}$ are orthogonal with respect to $w^{(a,b)}_P(x)$.

Assume that either of the following hold:
\begin{enumerate}
\item  \(\rho\) is a degree 1 polynomial, or 
\item \(\rho\) is the square root of a non-negative degree \(\le\) 2 polynomial, \(-\gamma = \delta > 0\), and \(w^{(a,b)}_P\) is an even function (i.e. $a = b$, and we can hence denote the weight as $w^{(a)}_P(x) = (\delta-x^2)^a$).
\end{enumerate}
Then the four-parameter 2D polynomials defined on $\Omega$ given by
\begin{align*}
	\hdopnk^{(a,b,c,d)}(x,y) := \genjacnmk^{(a, b, c+d+2k+1)}(x) \: \rho(x)^k \: P_k^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align*}
are orthogonal polynomials with respect to the weight 
\begin{align*}
	W^{(a,b,c,d)}(x,y) := \genjacw^{(a,b,c+d)}(x) \: w_P^{(d,c)}(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega. 
\end{align*}
\end{proposition}
\begin{proof}
See \cite[p55--56]{dunkl2014orthogonal}.
\end{proof}

For the half disk, the weight $W^{(a,b)}(x,y) = x^a \: (1-x^2-y^2)^b$ results from setting:
\begin{align*}
\begin{cases}
(\alpha,\beta) &:= (0.1) \\
(\gamma,\delta) &:= (-1,1) \\
\rho(x) &:= (1-x^2)^{\half} \\
w_1(x) &:= \genjacw^{(a,b)}(x) := x^a \: (1-x^2)^b \\
w_2(x) &:= w_P^{(b)}(x) := (1-x^2)^b = (1-x)^b \: (1+x)^b,
\end{cases}
\end{align*}
Note here the we have made the adjustment that $\genjacw^{(a,b,c)}(x) = (\beta - x)^a \: (x - \alpha)^{b} \: \rho(x)^{2c}$ and simply set the first parameter to zero and removed it for the specific half-disk case. This is for ease of writing and computation, by ensuring we reduce the need to incorporate unnecessary parameters and instead yield a two-parameter family of 2D orthogonal polynomials.

The weight $w_2(x)$ is an ultraspherical weight, and the corresponding OPs are   the Jacobi polynomials  \(\{P_n^{(b, b)}\}\). The weight $w_1(x)$ is non-classical (it is in fact semi-classical, and is equivalent to a generalized Jacobi weight \cite[\S5]{magnus1995painleve}) and we introduce the notation \(\{\genjac_n^{(a, b)}\}\) for the orthonormal polynomials. Thus we arrive at  a two parameter family of 2D orthogonal polynomials $\{\hdopnkab\}$ given by, for $0 \le k \le n, \: n = 0,1,2,\dots,$
\begin{align}\label{eq:diskpolys}
 \hdopnk^{(a,b)}(x,y) := \genjacnmk^{(a, b+k+\half)}(x) \: \rho(x)^k \: P_k^{(b,b)}\fpr(\frac{y}{\rho(x)}), \quad (x,y) \in \Omega, 
\end{align}
orthogonal on \(\Omega\) with respect to the weight
\begin{align*}
W^{(a,b)}(x,y) = \genjacw^{(a,b)}(x) \: w_P^{(b)}(\frac{y}{\rho(x)}) = x^a \: (1-x^2-y^2)^b.
\end{align*}

The two 1D families of orthogonal polynomials we have denoted $\{\genjac_{n}^{(a,b)}\}$ and $\{P_n^{(b,b)}\}$ are not orthonormal, but have (squared) norms given by the integral of their respective weight functions we denote as
\begin{align*}
	\normgenjac^{(a,b)} := \int_0^1 \: \genjacw^{(a,b)}(x) \: \D x, \quad \normjac^{(b)} := \int_{-1}^1 \: w_P^{(b)}(y) \: \D y,
\end{align*}
i.e. $\norm{\genjac_{n}^{(a,b)}}_{\genjacw^{(a,b)}}^2 = \normgenjac^{(a,b)}$ $\forall n = 0,1,2,\dots$ and $\norm{P_{n}^{(b,b)}}_{w_P^{(b)}}^2 = \normjac^{(b)}$ $\forall n = 0,1,2,\dots$.

@@

\begin{lemma}
$\hdopnk^{(a,b)}(x,y)$ satisfy the following 3-term recurrences:
\begin{align*}
x \hdopnk^{(a,b)}(x,y) &= \alphaab_{n,k,1} \: \hdop_{n-1, k}^{(a,b)}(x, y) + \alphaab_{n,k,2} \: \hdop_{n, k}^{(a,b)}(x, y) + \alphaab_{n+1,k,1} \: \hdop_{n+1, k}^{(a,b)}(x, y), \\
y \hdopnk^{(a,b)}(x,y) &= \betaab_{n,k,1} \: \hdop_{n-1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,2} \: \hdop_{n-1, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,3} \: \hdop_{n, k-1}^{(a,b)}(x, y) + \betaab_{n,k,4} \: \hdop_{n, k+1}^{(a,b)}(x, y) \nonumber \\
		& \quad \quad + \betaab_{n,k,5} \: \hdop_{n+1, k-1}^{(a,b)}(x, y) + \betaab_{n,k,6} \: \hdop_{n+1, k+1}^{(a,b)}(x, y),
\end{align*}
for \((x,y) \in \Omega\), where
\begin{align*}
\alphaab_{n,k,1} &:= \beta_{n-k-1}^{(a, b+k+\half)}, \qquad \alphaab_{n,k,2} := \alpha_{n-k}^{(a, b+k+\half)} \\
\betaab_{n,k,1} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,2} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-2}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,3} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+1}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,4} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k-1}^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}} \\
\betaab_{n,k,5} &:= \frac{\delta_{k-1}^{(b)}}{\normgenjac^{(a, b + k - \half)}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjac_{n-k+2}^{(a, b+k-1/2)}>_{\genjacw^{(a, b+k+1/2)}} \\
\betaab_{n,k,6} &:= \frac{\delta_{k}^{(b)}}{\normgenjac^{(a, b + k + \frac{3}{2})}} \: \ip<\genjacnmk^{(a, b+k+1/2)}, \genjacnmk^{(a, b+k+{3/2})}>_{\genjacw^{(a, b+k+{3/2})}}. 
\end{align*}

@@

\begin{proof}
First, note that, using a change of variable,
\begin{align*}
	\norm{\hdopnkab}_\Wab^2 &= \iint_\Omega \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \rho(x)^{2k} \: {P_k^{(b,b)}\fpr(\frac{y}{\rho(x)})^2} \: \genjacw^{(a,b)}(x) \: w_P^{(b)}\fpr(\frac{y}{\rho(x)}) \: \D y \: \D x \nonumber \\
	&= \Big( \int_\alpha^\beta \: {\genjacnmk^{(a,b+k+\half)}}(x)^2 \: \genjacw^{(a,b+k+\half)}(x) \: \D x \Big) \: \Big( \int_\gamma^\delta \: {P^{(b,b)}}(y)^2 \: w_P^{(b)}(y) \: \D y \Big) \nonumber \\
	&= \normgenjac^{(a,b+k+\half)} \: \normjac^{(b)}.
\end{align*}
where we use the notation that $\ip<p, q>_{\Wab} := \iint_\Omega p(x,y) \: q(x,y) \: \Wab(x,y) \: \D y \: \D x$ and $\norm{p}_{W^{(a,b)}}^2 := \ip<p, p>_{\Wab}$ .

Finally, note that \(\ip<y \hdopnkab, \hdopmjab>_\Wab = 0\) for \(m < n-1\) (and for \(m > n+2\)). Thus for \(m = n-1, n, n+1,\) \(j = 0,\dots,m:\)
\begin{align*}
\ip<y \hdopnkab, \hdopmjab>_\Wab &=  \iint_\Omega \hdopnkab(x,y) \: \hdopmjab(x,y) \: y \: \Wab(x,y) \: dy \: dx \\
&= \Big( \int^1_0 \genjacnmk^{(a, b+k+\half)}(s) \: \genjacmmj^{(a, b+j+\half)}(s) \: s^a \: \rho(s)^{2b+k+j+2} \: ds \Big) \nonumber \\
& \quad \quad \quad\cdot \: \Big( \int^1_{-1} P_k^{(b,b)}(t) \: P_j^{(b,b)}(t) \: t \: (1-t^2)^b \: dt \Big) \\
&= \begin{cases}
    	\delta_k^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k-1}^{(a, b+k+\frac{3}{2})}>_{\genjacw^{(a, b+k+\frac{3}{2})}} \quad& \text{if } j = k+1 \\
	\delta_{k-1}^{(b)} \: \normjac^{(b)} \: \ip<\genjacnmk^{(a, b+k+\half)}, \genjac_{m-k+1}^{(a, b+k-\half)}>_{\genjacw^{(a, b+k+\half)}} \quad& \text{if } j = k-1 \\
	0 & \text{otherwise}
      \end{cases}.
\end{align*}

@@

Three-term recurrences lead to Jacobi operators that correspond to multiplication by $x$ and $y$. Define, for $n=0,1,2,\dots$: 
\begin{align*}
\bighdopab_n := \begin{pmatrix}
		\hdopab_{n,0}(x,y) \\
		\vdots \\
		\hdopab_{n,n}(x,y)
	\end{pmatrix} \in \R^{n+1}, 
\quad \quad 
\bighdopab := \begin{pmatrix}
		\bighdopab_0 \\
		\bighdopab_1 \\
		\bighdopab_2 \\
		\vdots \\
	\end{pmatrix}
\end{align*}
and set $J_x^{(a,b)}, J_y^{(a,b)}$ as the Jacobi matrices corresponding to
\begin{align}
J_x^{(a,b)} \: \bighdopab(x,y) = x \: \bighdopab(x,y), \quad J_y^{(a,b)} \: \bighdopab(x,y) = y \: \bighdopab(x,y).
\label{eqn:jacobimatricesdefinition}
\end{align}
The matrices $J_x^{(a,b)}, J_y^{(a,b)}$ act on the coefficients vector of a functions expansion in the $\{\hdopnkab\}$ basis. For example, let $a, b$ be general parameters and a function $f(x,y)$ defined on $\Omega$ be approximated by its expansion $f(x,y) = \bighdopab(x,y)^\top \mathbf{f}$. Then $x \: f(x,y)$ is approximated by $\bighdopab(x,y)^\top {J_x^{(a,b)\top}} \mathbf{f}$. In other words, ${J_x^{(a,b)\top}} \mathbf{f}$ is the coefficients vector for the expansion of the function $(x,y) \to x \: f(x,y)$ in the  $\{\hdopnkab\}$ basis.
